Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-11-07 manuscript/manuscript.Rmd
minor revisions.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

@@

Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological handaxe assemblage known for its remarkable dexterity, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020]. This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010]. In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high dexerity reflected in their manufacture. As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019]. To identify the morphological manifestation of knappers' dexterity in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017]. The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].

@@

In this experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas from our expert sample. Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage. She was present at all training sessions to provide help and instruction to participants. All training occurred under controlled conditions at the outdoor knapping area of Emory's Paleolithic Technology Lab, with knapping tools and raw materials provided. All participants were instructed in basic knapping techniques including how to select appropriate percussors, initiate flaking on a nodule, maintain the correct flaking gestures and angles, prepare flake platforms, visualize outcomes, deal with raw material imperfections, and correct mistakes. Handaxe-specific instruction included establishment and maintenance of a bifacial plane, cross-sectional thinning, and overall shaping. The training emphasized both aspects of handaxe making technical skill (the importance of producing thin pieces with centered edges) as well as mental template related markers (symmetrical edges).

@@

In addition, visual inspection of the principle component scatter plot (Fig. ) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups. To test this, we conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)). Across all three groups, a negative correlation has been displayed between the PC1 and PC2 values, although this trend is not statistically significant (r=-0.41, p= 0.24) in the expert group, probably because of its small sample size.

>>>>

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

@@

Drawing on these two lines of literature, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020]. This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010]. In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture. As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019]. To identify the morphological manifestation of knappers' skill level in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017]. The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].

@@

In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas from our expert sample. Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage. She was present at all training sessions to provide help and instruction to participants. All training occurred under controlled conditions at the outdoor knapping area of Emory's Paleolithic Technology Lab, with knapping tools and raw materials provided. All participants were instructed in basic knapping techniques including how to select appropriate percussors, initiate flaking on a nodule, maintain the correct flaking gestures and angles, prepare flake platforms, visualize outcomes, deal with raw material imperfections, and correct mistakes. Handaxe-specific instruction included establishment and maintenance of a bifacial plane, cross-sectional thinning, and overall shaping. The training emphasized both aspects of handaxe making technical skill (the importance of producing thin pieces with centered edges) as well as mental template related markers (symmetrical edges).

@@

In addition, visual inspection of the principle component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups. To test this, we conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)). Across all three groups, a negative correlation has been displayed between the PC1 and PC2 values, although this trend is not statistically significant (r=-0.41, p= 0.24) in the expert group, probably because of its small sample size.
